Bayesian Hierarchical Modeling: An Introduction and Reassessment

### 2022-12-18 manuscript.Rmd
Updates revision

\Revision{In hierarchical analyses, the general or overall effect is sometimes called fixed effect, and the individual-specific deviations from this effect are in that case referred to as random effects (Rouder \& Lu, 2005). We can also estimate individual effects, not as deviations from an overarching effect, but as effects for each individual, as obtained from a per-participant analysis. Therefore, more common terms that suit the interpretation of the analyses better are general effect, individual deviation, and individual effects instead of fixed and random effects. We will use the terms general effect, individual deviation, and individual effect throughout this tutorial.}

@@

## Step 3: Prior Predictions 
A key advantage of model specification with priors is that we can obtain preditions on what the data would look like that are generated from the model. These prior predictions allow us to make the implications of our specified priors more concrete. Researchers can look at relevant statistic that summarize the simulated data and decide if they correspond to reasonable observations for their specific research context. If the simulated data are in line with the researchers' expectations, the priors can be used for the analysis. If the simulated data do not match the expectations, the priors have to be adjusted (Step 4).

@@

The normal distribution on $\mu_{\beta}$ and $\mu_{\delta}$ seems reasonable if we assume that a two-sided hypothesis about the symbolic distance effect is tested. However, we expect less variance around this mean. Therefore, we decrease the setting for the variation. The wide tails of the Student's *t*-distribution could be one of the reasons for the high range in variation. We change the prior for $\mu_{\gamma}$ to a normal distribution that has more narrow tails. As RTs cannot be negative, we adjust the prior to be positive only by truncating the normal distribution using a lower bound of zero. Finally, we would like to use a more common conjugate prior for the variances to ensure obtaining a proper posterior distribution. ^[Conjugate priors result in posterior distributions that belong to the same distribution family as the prior distribution [@Fink1997]. For a normal likelihood, these are the normal distribution for the mean and the inverse gamma distribution for the variance.] Therefore, we choose an inverse gamma distribution. We can also be more specific about the expected RTs. Based on our knowledge of these types of tasks [e.g., @haaf2017developing], it seems highly plausible that participants will respond within a second or two. We would expect effects in the range of 10 milliseconds to 100 milliseconds. We adjust our priors using these new distribution types and settings, evaluate the prior predictions (shown in panel B of Figure \@ref(fig:priorpredfigurepanel)) and arrive at the following priors:^[Because response times cannot be negative the prior for the overall baseline $\mu_\gamma$ is a truncated normal with a lower bound of zero.]

@@

The posterior distributions for the focal parameters in the normal model are presented in Figure \@ref(fig:modelposteriorsmufigure3). The posterior means estimated by `rstan` and `brms` can be found in [Online Supplement H](https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment/tree/main/H%20-%20Posterior%20Distributions%20Tables). The credible interval displayed in the figure is the Bayesian version of a confidence interval and provides a measure of uncertainty about the parameter value [@wagenmakers2018bayesian].

@@

For the digit effects, positive parameter estimates indicate that the response is slower for the non-baseline digits, that is, digits that are closer to 5. The parameters $\mu_{\delta_{6}}$ and $\mu_{\delta_{4}}$ are distributed around higher values and their credible intervals do not contain zero. Note that this result is consistent with the symbolic distance hypothesis that the digits 4 and 6 have the highest response times because they are closest to the 5, the comparison value. The 95% credible intervals of $\mu_{\delta_{7}}$ and $\mu_{\delta_{3}}$ contain zero. Therefore, it is plausible that the response times do not differ substantially from the baseline condition. The posterior distribution of the side effect $\mu_{\beta}$ is centered around zero with a posterior mean of `r printnum(fitrstanhiertable3["mu2", "Mean"], digits = 3)` and a 95% credible interval ranging from `r printnum(fitrstanhiertable3["mu2", "Lower Bound"], digits = 3)` to `r printnum(fitrstanhiertable3["mu2", "Upper Bound"], digits = 3)`. This indicates a small to non-existent side effect.

@@

Finally, Figure \@ref(fig:modelestimatesfigure3) shows how the relationship between individuals' digit effects.^[The design and code for this figure are by @j_van_langen_2020_3715576.] The posterior means of individual effects are represented by the dots. The right side of each plot represents the distribution of the individual estimates. Note that this distribution is not the posterior distribution of the general effect. It rather corresponds to the left side of Figure 6.

>>>>

\Revision{In hierarchical analyses, the general or overall effect is sometimes called fixed effect, and the individual-specific deviations from this effect are in that case referred to as random effects (Rouder \& Lu, 2005). We can also estimate individual effects, not as deviations from an overarching effect, but as effects for each individual, as obtained from a per-participant analysis. Throughout this tutorial, we will use the terms that suit the interpretation of our analyses, namely general effect, individual deviation, and individual effect.}

@@

## Step 3: Prior Predictions 
A key advantage of model specification with priors is that we can obtain preditions on what the data would look like that are generated from the model \Revision{[@etz2018bayesian; @lee2018determining]}. These prior predictions allow us to make the implications of our specified priors more concrete. Researchers can look at relevant statistic that summarize the simulated data and decide if they correspond to reasonable observations for their specific research context. If the simulated data are in line with the researchers' expectations, the priors can be used for the analysis. If the simulated data do not match the expectations, the priors have to be adjusted (Step 4).

@@

The normal distribution on $\mu_{\beta}$ and $\mu_{\delta}$ seems reasonable if we assume that a two-sided hypothesis about the symbolic distance effect is tested. However, we expect less variance around this mean. Therefore, we decrease the setting for the variation. The wide tails of the Student's *t*-distribution could be one of the reasons for the high range in variation. We change the prior for $\mu_{\gamma}$ to a normal distribution that has more narrow tails. As RTs cannot be negative, we adjust the prior to be positive only by truncating the normal distribution using a lower bound of zero. Finally, we would like to use a more common conjugate prior for the variances to ensure obtaining a proper posterior distribution. ^[Conjugate priors result in posterior distributions that belong to the same distribution family as the prior distribution [@Fink1997]. For a normal likelihood, these are the normal distribution for the mean and the inverse gamma distribution for the variance.]\Revision{However, in `rstan`, non-conjugate priors are commonly applied [e.g., @gelman2006prior]} Therefore, we choose an inverse gamma distribution. We can also be more specific about the expected RTs. Based on our knowledge of these types of tasks [e.g., @haaf2017developing], it seems highly plausible that participants will respond within a second or two. We would expect effects in the range of 10 milliseconds to 100 milliseconds. We adjust our priors using these new distribution types and settings, evaluate the prior predictions (shown in panel B of Figure \@ref(fig:priorpredfigurepanel)) and arrive at the following priors:^[Because response times cannot be negative the prior for the overall baseline $\mu_\gamma$ is a truncated normal with a lower bound of zero.]

@@

The posterior distributions for the focal parameters in the normal model are presented in Figure \@ref(fig:modelposteriorsmufigure3). The posterior means estimated by `rstan` and `brms` can be found in [Online Supplement H](https://github.com/MyrtheV/Bayesian-Hierarchical-Modelling-An-Introduction-and-Reassessment/tree/main/H%20-%20Posterior%20Distributions%20Tables). The credible interval displayed in the figure is the Bayesian version of a confidence interval and provides a measure of uncertainty about the parameter value [@wagenmakers2018bayesian]. \Revision{"Without considering the prior, a posterior credible interval does not directly allow one to conclude that there is evidence for or against an effect." It merely tells us }

@@

For the digit effects, positive parameter estimates indicate that the response is slower for the non-baseline digits, that is, digits that are closer to 5. The parameters $\mu_{\delta_{6}}$ and $\mu_{\delta_{4}}$ are distributed around higher values and their credible intervals do not contain zero. Note that this result is consistent with the symbolic distance hypothesis that the digits 4 and 6 have the highest response times because they are closest to the 5, the comparison value. The 95% credible intervals of $\mu_{\delta_{7}}$ and $\mu_{\delta_{3}}$ contain zero. \Revision{Therefore, it is plausible that the response times do not differ substantially from the baseline condition.- is this one necessary to adjust?} The posterior distribution of the side effect $\mu_{\beta}$ is centered around zero with a posterior mean of `r printnum(fitrstanhiertable3["mu2", "Mean"], digits = 3)` and a 95% credible interval ranging from `r printnum(fitrstanhiertable3["mu2", "Lower Bound"], digits = 3)` to `r printnum(fitrstanhiertable3["mu2", "Upper Bound"], digits = 3)`. \Revision{This indicates a small to non-existent side effect. -change}

@@

\Revision{The individual variation can also be visualized in a different way with focusing on the relation between the individual estimates.} Figure \@ref(fig:modelestimatesfigure3) shows how the relationship between individuals' digit effects.^[The design and code for this figure are by @j_van_langen_2020_3715576.] The posterior means of individual effects are represented by the dots. The right side of each plot represents the distribution of the individual estimates. Note that this distribution is not the posterior distribution of the general effect. It rather corresponds to the left side of Figure 6.

### 2022-12-18 manuscript.tex
Updates revision

The application of Bayesian hierarchical analysis will be illustrated using a digit classification task collected by Rouder, Lu, Speckman, Sun, and Jiang (2005). In an experimental task participants repeatedly indicated whether a target digit (i.e., 2, 3, 4, 6, 7, or 8) was greater or smaller than 5. The participants in the task performed multiple trials per condition (i.e., digit). Therefore, observations are nested in participants. Our analyses focus on two effects that can potentially occur in this digit classification task: The digit effect and the side effect. According to the symbolic distance hypothesis (Moyer \& Landauer, 1967) processing of numbers is analogue. If numbers are closer on the number line, we may confuse them more, and comparing them is more difficult. In line with this hypothesis, the digit effect postulates that response times (RTs) are slower when digits are closer to 5. Additionally, we may hypothesize that RTs are affected by whether the target digit was smaller or greater than 5, yet it is unclear which side of 5 would lead to faster or slower RTs. \textcolor{blue}{The digit classification task is a standard task that usually would be analyzed in a 3 (digit, closest to 5, further away from 5, furthest away from 5) by 2 (side, greater or samller than 5) within-subjects ANOVA.} In this tutorial, we show how inferences can be improved using a hierarchical Bayesian model instead.

@@

Finally, Figure \ref{fig:modelestimatesfigure3} shows how the relationship between individuals' digit effects.\footnote{The design and code for this figure are by Langen (2020).} The posterior means of individual effects are represented by the dots. The right side of each plot represents the distribution of the individual estimates. Note that this distribution is not the posterior distribution of the general effect. It rather corresponds to the left side of Figure 6.

>>>>

The application of Bayesian hierarchical analysis will be illustrated using a digit classification task collected by Rouder, Lu, Speckman, Sun, and Jiang (2005). In an experimental task participants repeatedly indicated whether a target digit (i.e., 2, 3, 4, 6, 7, or 8) was greater or smaller than 5. The participants in the task performed multiple trials per condition (i.e., digit). Therefore, observations are nested in participants. Our analyses focus on two effects that can potentially occur in this digit classification task: The digit effect and the side effect. According to the symbolic distance hypothesis (Moyer \& Landauer, 1967) processing of numbers is analogue. If numbers are closer on the number line, we may confuse them more, and comparing them is more difficult. In line with this hypothesis, the digit effect postulates that response times (RTs) are slower when digits are closer to 5. Additionally, we may hypothesize that RTs are affected by whether the target digit was smaller or greater than 5, yet it is unclear which side of 5 would lead to faster or slower RTs. \textcolor{blue}{The digit classification task is a standard task that usually would be analyzed in a 3 (digit, closest to 5, further away from 5, furthest away from 5) by 2 (side, greater or smaller than 5) within-subjects ANOVA.} In this tutorial, we show how inferences can be improved using a hierarchical Bayesian model instead.

@@

The individual variation can also be visualized in slightly different way, while focusing on the relation between the individual estimates. Figure \ref{fig:modelestimatesfigure3} shows how the relationship between individuals' digit effects.\footnote{The design and code for this figure are by Langen (2020).} The posterior means of individual effects are represented by the dots. The right side of each plot represents the distribution of the individual estimates. Note that this distribution is not the posterior distribution of the general effect. It rather corresponds to the left side of Figure 6.
